Shapes analysis for time series.

### 2024-05-07 NEURIPS24/neurips_2024_to_submit.tex
update

\begin{abstract}
    [TO CHANGE]
    Analyzing the inter-individual variability within a cluster of time series is particularly appealing in medicine and biology.
    Individuals have specificities depending on their genes or disease
     while sharing, on average, a similar pattern in the features of their corresponding time series. For instance, an electrocardiogram presents a typical shape of the heartbeat, which varies depending on the individual.
    In this paper, we propose an unsupervised representation method (TS-LDDMM) to analyze inter-individual variability of time series dataset.
    Inspired by the shape analysis literature, we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) to time series, considering the temporal evolution as their distinctive feature.  
    % Given a dataset of $N$ individual's time series $(s^j)_{j\in[N]}$,
    %  we learn a time series of reference $\mathbf{s}_0$ encoding the common part of the individuals as well as diffeomorphisms $(\phi^j)_{j\in[N]}$ encoding the specificities of each individual:
    %  $s^j$ is seen as the transformation of $\mathbf{s}_0$ by $\phi_j$.
    %  Then, the parameters $(\alpha_j)_{j\in[N]}$ encoding the diffeomorphisms $(\phi^j)_{j\in[N]}$ are used as the features representation related to each time series $s^j$.
    In this paper, we propose an unsupervised representation learning method (URL) for time series \cite{trirat2024universal,meng2023unsupervised} by giving a special attention to their \textit{shapes} for studying inter-variability in biomedical applications.
    In particular, the method handle time series with \textbf{irregular sampling} and \textbf{different sizes}.
    This work belongs to the shape analysis literrature: %which focus on the statistical analysis of various mathematical objects invariant under rotations, dilations, or time parameterization.
     we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) \cite{beg2005computing,vaillant2004statistics} to the case of time series data by applying difformations to their graph.
    However, time series graph have more structure than 1D curves:
     a graph time series transformation by a general diffeomorphism is not always a graph time series. Our contribution arise from this observation.
    This work is at the crossroad of URL for time series and shape analysis; we hope it will benefit to both communities.
    % These features can be post-processed with any statistical or machine learning tools. 
     We demonstrate the advantages of our representation compared to existing methods using synthetic data and real-world examples, motivated by applications in medicine.
\end{abstract}

>>>>

\begin{abstract}
    %TO CHANGE]
    %Analyzing the inter-individual variability within a cluster of time series is particularly appealing in medicine and biology.
    %Individuals have specificities depending on their genes or disease
     %while sharing, on average, a similar pattern in the features of their corresponding time series. For instance, an electrocardiogram presents a typical shape of the heartbeat, which varies depending on the individual.
    %In this paper, we propose an unsupervised representation method (TS-LDDMM) to analyze inter-individual variability of time series dataset.
    %Inspired by the shape analysis literature, we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) to time series, considering the temporal evolution as their distinctive feature.  
    % Given a dataset of $N$ individual's time series $(s^j)_{j\in[N]}$,
    %  we learn a time series of reference $\mathbf{s}_0$ encoding the common part of the individuals as well as diffeomorphisms $(\phi^j)_{j\in[N]}$ encoding the specificities of each individual:
    %  $s^j$ is seen as the transformation of $\mathbf{s}_0$ by $\phi_j$.
    %  Then, the parameters $(\alpha_j)_{j\in[N]}$ encoding the diffeomorphisms $(\phi^j)_{j\in[N]}$ are used as the features representation related to each time series $s^j$.
    In this paper, we propose an unsupervised representation learning method (URL) for time series %\cite{trirat2024universal,meng2023unsupervised}
     by giving a special attention to their \textit{shapes} for studying inter-variability in biomedical applications.
    In particular, the method can handle \textit{irregularly sampled} time series having \textit{different sizes}.
    This work belongs to the shape analysis literrature: %which focus on the statistical analysis of various mathematical objects invariant under rotations, dilations, or time parameterization.
     we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) %\cite{beg2005computing,vaillant2004statistics}
      to the case of time series data by applying deformations to their graph.
    However, the method is not a simple application of LDDMM since time series graph have more structure than 1D curves.
     Indeed, a graph time series transformed by a general diffeomorphism is not always a graph time series.
      We solve this issue by establishing a representation theorem on time series graph and derive its consequences on LDDMM.
    This work is at the crossroad of URL for time series and shape analysis; we hope it will benefit to both communities.
    % These features can be post-processed with any statistical or machine learning tools. 
     We demonstrate the advantages of our representation compared to existing methods using synthetic data and real-world examples, motivated by applications in medicine.
\end{abstract}

### 2024-05-07 NEURIPS24/paper.tex
update

In this paper, we propose an unsupervised representation learning method (URL) for time series \cite{trirat2024universal,meng2023unsupervised} by giving a special attention to their \textit{shapes} for studying inter-variability in biomedical applications.
In particular, the method handle time series with \textbf{irregular sampling} and \textbf{different sizes}.
This work belongs to the shape analysis literrature: %which focus on the statistical analysis of various mathematical objects invariant under rotations, dilations, or time parameterization.
 we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) \cite{beg2005computing,vaillant2004statistics} to the case of time series data by applying difformations to their graph.
However, time series graph have more structure than 1D curves:
 a graph time series transformation by a general diffeomorphism is not always a graph time series.
This work is at the crossroad of URL for time series and shape analysis; we hope it will benefit to both communities.
%A time series is a set of values indexed by time \cite{trirat2024universal}.
%This type of data is prevalent in biology and medicine for recording physiological time series , as well as in the broader context of the Internet of Things sensors \cite{fathy2018large} and measurements in physical systems \cite{luo2021deep}.
 %However, due to the nature of such data, which have different sample sizes and incorporate a temporal evolution, classical statistical analysis, and machine learning methods need refinement \cite{anderson2011statistical}.
 %One option to tackle this issue is to derive a feature representation of time series, which depends on the problem at hand \cite{trirat2024universal}.

@@

  Although a community structure with representatives can be learned in an unsupervised way using contrastive loss \cite{franceschi2019unsupervised,tonekaboni2021unsupervised,meng2023unsupervised}
  or similarity measures \cite{asgari2023clustering,germain2023unsupervised,paparrizos2015k,ye2009time},
     studying the inter-individual variability of shapes within a cluster \cite{niennattrakul2007inaccuracies,shirato2023identifying} is still an open problem in URL.
     %In this paper, we aim to go beyond group structure by giving an unsupervised data-driven method that represents the specificity of each individual's shape \cite{vidal2021perceptual}.
  %For instance, the shape of a heartbeat in an Electrocardiogram (ECG) -the pattern- has been well studied \cite{madona2021pqrst} with the PQRST decomposition (see figure\sam{fill}).
  %There exist several methods to detect and segment patterns in a time series (\sam{mettre ref thibuat} ).
   %The goal is then to study their shape according to the property of the individual.

@@

 First, we propose not to see time series through their curve $\{s_t:\eqsp t\in\msi\}$, but through their graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\} $. 
 Then, building on the shape analysis litterature, we follow the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs.
  The idea is to represent each element $(\msg(s^j))_{j\in[N]}$ of the dataset as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism.
  Then, the diffeomorphism is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS).
   The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$.
    Finally, these features encoding the shapes can feed any statistical or machine learning model as in URL.

   However, a graph time series transformation by a general diffeomorphism is not always a graph time series, see e.g. \Cref{fig:diffeo}, thus a graph time series is more than a simple 1D curve \cite{glaunes2008large}.
   Our contributions arise from this observation.
   %In this regard, this work is not an application of LDDMM to 1D curves \cite{glaunes2008large}.
    We solve this issue by specifying the class of diffeomorphisms to consider and showing how to learn them.
     This change is fruitful in representing time transformation as illustrated in \Cref{fig:transport}.

>>>>

% In this paper, we propose an unsupervised representation learning method (URL) for time series \cite{trirat2024universal,meng2023unsupervised} by giving a special attention to their \textit{shapes} for studying inter-variability in biomedical applications.
% In particular, the method handle time series with \textbf{irregular sampling} and \textbf{different sizes}.
% This work belongs to the shape analysis literrature: %which focus on the statistical analysis of various mathematical objects invariant under rotations, dilations, or time parameterization.
%  we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) \cite{beg2005computing,vaillant2004statistics} to the case of time series data by applying difformations to their graph.
% However, time series graph have more structure than 1D curves:
%  a graph time series transformation by a general diffeomorphism is not always a graph time series.
% This work is at the crossroad of URL for time series and shape analysis; we hope it will benefit to both communities.
%A time series is a set of values indexed by time \cite{trirat2024universal}.
%This type of data is prevalent in biology and medicine for recording physiological time series , as well as in the broader context of the Internet of Things sensors \cite{fathy2018large} and measurements in physical systems \cite{luo2021deep}.
 %However, due to the nature of such data, which have different sample sizes and incorporate a temporal evolution, classical statistical analysis, and machine learning methods need refinement \cite{anderson2011statistical}.
 %One option to tackle this issue is to derive a feature representation of time series, which depends on the problem at hand \cite{trirat2024universal}.

@@

  Although a community structure with representatives can be learned in an unsupervised way \cite{trirat2024universal,meng2023unsupervised} using contrastive loss \cite{franceschi2019unsupervised,tonekaboni2021unsupervised,meng2023unsupervised}
  or similarity measures \cite{asgari2023clustering,germain2023unsupervised,paparrizos2015k,ye2009time},
     studying the inter-individual variability of shapes within a cluster \cite{niennattrakul2007inaccuracies,shirato2023identifying} is still an open problem in URL.
     %In this paper, we aim to go beyond group structure by giving an unsupervised data-driven method that represents the specificity of each individual's shape \cite{vidal2021perceptual}.
  %For instance, the shape of a heartbeat in an Electrocardiogram (ECG) -the pattern- has been well studied \cite{madona2021pqrst} with the PQRST decomposition (see figure\sam{fill}).
  %There exist several methods to detect and segment patterns in a time series (\sam{mettre ref thibuat} ).
   %The goal is then to study their shape according to the property of the individual.

@@

 First, we propose not to see time series through their curve $\{s_t:\eqsp t\in\msi\}$, but through their graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\} $. 
 Then, building on the shape analysis litterature \cite{beg2005computing,vaillant2004statistics}, we follow the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs.
  The idea is to represent each element $(\msg(s^j))_{j\in[N]}$ of the dataset as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism.
  Then, the diffeomorphism is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS).
   The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$.
    Finally, these features encoding the shapes can feed any statistical or machine learning model as in URL.

   However, a graph time series transformation by a general diffeomorphism is not always a graph time series, see e.g. \Cref{fig:diffeo}, thus a graph time series is more than a simple 1D curve \cite{glaunes2008large}.
   Our contributions arise from this observation.
   %In this regard, this work is not an application of LDDMM to 1D curves \cite{glaunes2008large}.
    We solve this issue by specifying the class of diffeomorphisms to consider and showing how to learn them.
     This change is fruitful in representing time transformation as illustrated in \Cref{fig:transport}.
     In particular, the method can handle \textit{irregularly sampled} time series having \textit{different sizes}.
